# Reproducible Research: Peer Assessment 1
Created by Sridher Kaminani on September 14, 2014

### Basic settings to Make the code visible
```{r}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
```


## Loading and preprocessing the data
```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))
cleanData <- na.omit(data)
rownames(cleanData) <- 1:nrow(cleanData)
head(cleanData)
dim(cleanData)
```


## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day
```{r}
library(ggplot2)
ggplot(cleanData, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```

2. Calculate and report the mean and median total number of steps taken per day

Mean total number of steps taken per day:
```{r}
totalSteps <- aggregate(cleanData$steps, list(Date = cleanData$date), FUN = "sum")$x
mean(totalSteps)
```

Median total number of steps taken per day:
```{r}
median(totalSteps)
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgSteps <- aggregate(cleanData$steps, list(interval = as.numeric(as.character(cleanData$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "steelblue", size = 0.8) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
avgSteps [avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```


## Inputing missing values

1. The total number of rows with NAs:

```{r}
sum(is.na(data))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

   My strategy is to use the mean for that 5-minute interval to fill each NA value in the steps column.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
newData <- data 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}

head(newData)
sum(is.na(newData))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
ggplot(newData, aes(date, steps)) + geom_bar(stat = "identity",
                                             colour = "steelblue",
                                             fill = "steelblue",
                                             width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")
```

## Are there differences in activity patterns between weekdays and weekends?

Mean total number of steps taken per day:
```{r}
newTotalSteps <- aggregate(newData$steps, 
                           list(Date = newData$date), 
                           FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
```
Median total number of steps taken per day:
```{r}
newMedian <- median(newTotalSteps)
newMedian
```
Compare them with the two before imputing missing data:
```{r}
oldMean <- mean(totalSteps)
oldMedian <- median(totalSteps)
newMean - oldMean
newMedian - oldMedian
```
So, after imputing the missing data, the new mean of total steps taken per day is the same as that of the old mean; the new median of total steps taken per day is greater than that of the old median.

### Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
head(newData)
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday","Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newData$weekdays)
table(newData$weekdays)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice)

avgSteps <- aggregate(newData$steps, 
                      list(interval = as.numeric(as.character(newData$interval)), 
                           weekdays = newData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"

xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```
 End of file
